---
title: "Apps User Guide"
subtitle: "Step-by-step walkthrough of the Scores and Species applications"
---

This guide walks through the two interactive applications — **Scores**
([app.marinesensitivity.org/scores](https://app.marinesensitivity.org/scores/))
and **Species**
([app.marinesensitivity.org/species](https://app.marinesensitivity.org/species/))
— one control at a time. Each numbered step below mirrors a step in that app's
built-in **Take a Tour** (click *About → Take a Tour* in either app to walk the
same path live), with the relevant control highlighted on the screenshot.

For a feature-by-feature reference instead, see the
[Scores](apps/scores.qmd) and [Species](apps/species.qmd) appendix pages.

::: callout-tip
## How this guide stays in sync

The steps, captions, and highlighted screenshots are generated directly from
the apps' in-app tours (the `Conductor$new()$step(...)` chains in each
`app.R`). Running `libs/apps-guide-regenerate.R` re-parses both tours,
re-highlights the same CSS element each tour points at, and re-captures the
screenshots, writing `figures/apps-guide/steps.json`. So whenever a tour
changes, re-running that one script refreshes this whole page.
:::

```{r}
#| label: setup
#| include: false
librarian::shelf(jsonlite, here, quiet = TRUE)

manifest <- here::here("figures", "apps-guide", "steps.json")
steps    <- if (file.exists(manifest)) jsonlite::read_json(manifest) else list()

# staleness check: is a tour source newer than the generated manifest?
apps_dir  <- Sys.getenv(
  "MSENS_APPS_DIR",
  unset = normalizePath(file.path(here::here(), "..", "apps"), mustWork = FALSE))
app_files <- file.path(apps_dir, c("mapgl", "mapsp"), "app.R")
stale     <- file.exists(manifest) &&
  any(file.exists(app_files) & file.mtime(app_files) > file.mtime(manifest))
```

```{r}
#| label: guide
#| output: asis
if (length(steps) == 0) {
  cat(
    "::: callout-warning\n",
    "No screenshot manifest found. Run ",
    "`Rscript libs/apps-guide-regenerate.R` (needs the sibling `apps` repo ",
    "and the live apps) to generate `figures/apps-guide/steps.json` and the ",
    "annotated screenshots.\n",
    ":::\n", sep = "")
} else {
  if (isTRUE(stale)) cat(
    "::: callout-warning\n",
    "A tour in `app.R` is newer than these screenshots — re-run ",
    "`libs/apps-guide-regenerate.R` to refresh this page.\n",
    ":::\n\n", sep = "")

  for (app in unique(vapply(steps, `[[`, "", "app"))) {
    cat(sprintf("\n## %s app\n\n", app))
    for (s in Filter(function(x) x$app == app, steps)) {
      cat(sprintf("\n### %d. %s\n\n", s$n, s$title))
      cat(s$text, "\n\n")
      cat(sprintf(
        "![%s](%s){#fig-%s-%02d fig-alt=\"%s\"}\n\n",
        s$title, s$image, s$app_key, s$n, gsub('"', "'", s$text)))
    }
  }
}
```
