# Extinction Risk {#sec-extinction-risk}

```{r}
#| include: false
library(dplyr)
tcols   <- doc_cols("taxon")$column_name
has_er  <- "er_score" %in% tcols
k       <- .doc_keys()
valid   <- paste0(k$valid, if (!is.na(k$marine)) paste0(" AND ", k$marine) else "")
```

Extinction-risk scoring turns a species' conservation status into a numeric weight, so that a cell holding rare or threatened species scores higher than one holding common ones. The MST combines three US federal frameworks with the international IUCN Red List into a single score from 1 to 100.

```{r}
#| label: er-absent
#| output: asis

if (!has_er) cat(
  "::: {.callout-warning}\n## Not published by ", doc_ver(), "\n\n",
  "This release's `taxon` table carries no `er_score` column, so the per-species weights ",
  "below cannot be shown for it. The framework is described; the counts are not, because ",
  "this release does not publish them.\n:::\n\n", sep = "")
```

## Regulatory framework

### Endangered Species Act (ESA)

The [Endangered Species Act](https://www.fws.gov/law/endangered-species-act) (1973) protects species listed as **Endangered (EN)** or **Threatened (TN)** by NMFS (marine species) or FWS (terrestrial and coastal species). ESA listing is the strongest signal available and drives the score for US-listed species.

### Marine Mammal Protection Act (MMPA) — assigned by taxonomy

The [Marine Mammal Protection Act](https://www.fisheries.noaa.gov/topic/laws-policies#marine-mammal-protection-act) (1972) protects **all** marine mammals in US waters regardless of ESA status, so the MST assigns it **by taxonomy**: every taxon in WoRMS class **Mammalia**, matched on AphiaID.

This is deliberate. Deriving the flag from the NMFS Species Directory instead — as earlier releases did — captures only the managed stocks listed there, roughly 70 against the ~137 marine-mammal taxa the analysis actually carries, and it silently omitted the FWS-managed species entirely (walrus, polar bear, sea otter, manatee, dugong). Taxonomy is the complete test because the statute is written in taxonomic terms.

The floor is **20**, higher than the MBTA's, because the MMPA imposes a stronger statutory standard: it affirmatively requires marine mammal populations to be maintained at Optimum Sustainable Population levels and mandates recovery, whereas the MBTA is primarily a prohibitive take statute. Marine mammals also tend to have lower reproductive rates and longer recovery times than most seabirds.

### Migratory Bird Treaty Act (MBTA) — an explicit species list, not all birds

The [Migratory Bird Treaty Act](https://www.fws.gov/law/migratory-bird-treaty-act-1918) (1918) protects **migratory birds native to the United States and its territories** — which is an explicit list, not a taxonomic class.

::: {.callout-important}
## The MBTA floor is the FWS **CFR 50 §10.13** list, not all Aves

A bird absent from 10.13 — a non-US or non-migratory species — receives **no** MBTA floor. Treating "bird" as equivalent to "MBTA-protected" would apply a US statutory floor to species the statute does not cover, inflating their weight in US waters.

Note also that most 10.13 birds are terrestrial. The list sets the *value* of the birds that reach scoring; what keeps terrestrial birds out of the score in the first place is the marine-relevance cull (see @sec-is-marine).
:::

### IUCN Red List

For species with no US federal listing, the score falls back to the [IUCN Red List](https://www.iucnredlist.org) category: Critically Endangered (CR), Endangered (EN), Vulnerable (VU), Near Threatened (NT), Least Concern (LC) and Data Deficient (DD).

::: {.callout-note}
`TN` is an **ESA** status (Threatened), not an IUCN one — IUCN uses `NT` for Near Threatened. `msens::compute_er_score()` **errors** on a code such as `IUCN:TN` rather than scoring it through a default, because a silently-defaulted code is a wrong weight that nothing surfaces.
:::

## Scoring

Scores are computed by `msens::compute_er_score()`, on two tracks.

### US-listed species

For a species with an ESA listing, or MMPA or MBTA protection, the score is the **maximum** of the applicable components:

$$
\text{er\_score} = \max(\text{ESA base},\; \text{MMPA floor},\; \text{MBTA floor})
$$ {#eq-er-us}

| Component | Score |
|:----------|------:|
| ESA Endangered (EN) | 100 |
| ESA Threatened (TN) | 50 |
| ESA not listed (LC) | 1 |
| MMPA protected (all Mammalia) | 20 |
| MBTA protected (CFR 50 §10.13) | 10 |

: Components for a US-listed species; the final score is the maximum across those that apply. {#tbl-er-us}

Worked examples:

- ESA Endangered whale (also MMPA): $\max(100, 20) = \mathbf{100}$
- ESA Threatened seabird on the 10.13 list: $\max(50, 10) = \mathbf{50}$
- Marine mammal with no ESA listing: $\max(1, 20) = \mathbf{20}$
- 10.13 migratory bird with no ESA listing: $\max(1, 10) = \mathbf{10}$

A species carrying an MMPA or MBTA floor is routed onto this track **even when it has no ESA listing**, which is what guarantees the floor is actually applied rather than being lost to the IUCN track.

### Species with no US listing

| IUCN category | Score |
|:--------------|------:|
| Critically Endangered (CR) | 50 |
| Endangered (EN) | 25 |
| Vulnerable (VU) | 5 |
| Near Threatened (NT) | 2 |
| Least Concern / Data Deficient / other | 1 |

: IUCN-based scores for species not listed by NMFS or FWS. {#tbl-er-iucn}

The IUCN scale is deliberately lower than the ESA one: a US statutory listing is a stronger, jurisdiction-specific statement about a species in US waters than an international assessment.

Species with no matching code at all receive the baseline of **1**.

### The governing score, and where it is applied

A taxon may carry several conservation codes at once — one per contributing dataset — and they can disagree, or be stale. The pipeline therefore computes one **governing** score per taxon: the most protective across its datasets, with a **US national listing overriding the IUCN category**, plus the statutory floors.

That governing score is applied **at merge time**, to the range cells — the point at which a species' presence in a cell should carry its governing risk rather than whatever code the source dataset happened to record. Scoring then reads values that already reflect it.

### When the listing is not the whole species: spatial extinction risk

A single governing score assumes the species is listed as a species. Under the ESA the *listed entity* can instead be a Distinct Population Segment, an ESU or a subspecies, and for some species only certain populations are listed: the humpback whale has six listed DPSs and eight that are not at risk — including the ones that breed in US waters — so its species page, flattened to the highest domestic status, read Endangered everywhere. For every species whose listed entities are scoped below the species (the list is in @sec-data-sources under *NMFS Distinct Population Segments*), risk is therefore a **surface**, not a scalar: each listed entity's designated critical habitat, from NOAA Fisheries' authoritative service, carries that entity's status (Endangered 100 / Threatened 50), and the rest of the species' IUCN range carries the baseline every unlisted species of its kind gets — the MMPA floor of 20 for a marine mammal, the IUCN category otherwise. Sea turtles have long had the same treatment through their own SWOT + DPS dataset. How that surface reaches the score differs between the two groups and is described in @sec-model-merging.

```{r}
#| label: tbl-er-dist
#| tbl-cap: !expr paste0("Distribution of the governing `er_score` across ", doc_ver(), "'s valid species.")

if (has_er) {
  d <- doc_tbl("taxon", sprintf(
    "SELECT er_score, count(*) AS n FROM taxon WHERE %s AND er_score IS NOT NULL
      GROUP BY 1 ORDER BY 1 DESC", valid))
  knitr::kable(
    data.frame(`er_score` = d$er_score,
               Species    = n_fmt(d$n),
               Share      = sprintf("%.1f%%", 100 * d$n / sum(d$n)),
               check.names = FALSE, row.names = NULL),
    align = c("r", "r", "r"))
}
```

```{r}
#| label: er-floors
#| output: asis

if (has_er && all(c("is_mmpa", "is_mbta") %in% tcols)) {
  d <- doc_tbl("taxon", sprintf(
    "SELECT sum(CASE WHEN is_mmpa THEN 1 ELSE 0 END) AS mmpa,
            sum(CASE WHEN is_mbta THEN 1 ELSE 0 END) AS mbta,
            count(*) AS n FROM taxon WHERE %s", valid))
  cat("Of ", doc_ver(), "'s ", n_fmt(d$n), " valid species, **", n_fmt(d$mmpa),
      "** carry the MMPA floor and **", n_fmt(d$mbta), "** the MBTA floor.\n\n", sep = "")
}
```

## Scale

Scores run from 1 to 100 and represent a percentage-like weight — 100 being maximum risk and 1 the baseline. Integers are used for storage efficiency; conceptually the scale is a 0–1 weight multiplied by 100.

## Implementation

```r
msens::compute_er_score(
  extrisk_code,      # "NMFS:EN", "FWS:TN", "IUCN:CR", ...
  is_mmpa = FALSE,   # MMPA protection (all WoRMS Mammalia)
  is_mbta = FALSE)   # MBTA protection (FWS CFR 50 §10.13)
```

`extrisk_code` is `AUTHORITY:STATUS`, where the authority is `NMFS`, `FWS` or `IUCN`, and the status is an ESA code (`EN`, `TN`, `LC`) or an IUCN code (`CR`, `EN`, `VU`, `NT`, `LC`, `DD`). Any other combination is an error, not a default.
