# Preface {.unnumbered}

```{r}
#| include: false
# knitr evaluates inline `r ...` only in a document that has at least one code
# chunk. Without this, the doc_*() calls below render as LITERAL TEXT — which is
# exactly what happened when the old stats.R setup chunk was removed.
```


The **Marine Sensitivity Toolkit** (MST) is a cloud-native system developed for the Bureau of Ocean Energy Management (BOEM) to assess the sensitivity of marine ecosystems to offshore energy development.

```{r}
#| label: index-headline
#| output: asis
cat("It integrates ", doc_stat("valid_species"),
    " species distribution models with extinction risk assessments across US waters", sep = "")
if (doc_can("programareas")) cat(" (", doc_stat("species_program_areas"),
    " of which fall within the ", doc_stat("n_program_areas"),
    " BOEM Program Areas of the current program cycle)", sep = "")
cat(".\n\n")
```

```{r}
#| label: index-changes
#| output: asis
# The first thing a reader of a versioned book needs is what THIS version is —
# standardized headings so two releases can be compared line for line.
doc_changes_callout()
```



![Overview of the MST methodology: species distribution models from the source datasets are merged, weighted by extinction risk scores (incorporating ESA, MMPA, MBTA, and IUCN protections), rescaled by ecoregion, and aggregated to management zones for visualization.](figures/overview-methods.svg){#fig-overview-preface}

### Interactive Apps

- [**Composite Scores**](https://app.marinesensitivity.org/scores/) — sensitivity scores by zone, with flower-plot summaries
- [**Species Distribution**](https://app.marinesensitivity.org/species/) — individual species distribution models and habitat suitability

Both render **any published release**: append `?ver=v3` (or any version) to view that one.
- [**Scoring Walkthrough**](scoring-walkthrough.qmd) — step through the scoring pipeline interactively, from input rasters to final scores

### Documentation

This book documents the scientific methodology, software infrastructure, and interactive applications that comprise the MST. It is organized in two parts:

- **@sec-science** — data sources, taxonomic integration, extinction risk scoring, model merging, and sensitivity scoring methodology
- **@sec-software** — server architecture, databases, workflows, APIs, R libraries, and interactive applications

### How to Cite

> Best, B.D., T.P. White (2026). Marine Sensitivity Toolkit: Project Documentation. Bureau of Ocean Energy Management, Environmental Studies Program. <https://marinesensitivity.org/docs/>

### Acknowledgments

This work is funded by the [BOEM Environmental Studies Program](https://www.boem.gov/environment/environmental-studies/environmental-studies-information) (BOEM-ESP). We thank the data providers whose open datasets make this work possible: [AquaMaps](https://www.aquamaps.org), [BirdLife International](https://www.birdlife.org), [IUCN Red List](https://www.iucnredlist.org), [NOAA Fisheries](https://www.fisheries.noaa.gov), and [U.S. Fish & Wildlife Service](https://www.fws.gov).
