# Model Merging {#sec-model-merging}

```{r}
#| include: false
library(dplyr)
mdl_key_style <- doc_manifest()$id_field == "mdl_key"
```

A taxon may have distribution data in several source datasets. Merging combines them into **one surface per taxon**, applying the taxon's governing extinction risk and constraining the result to where experts say the species actually occurs.

::: {.callout-tip}
## Interactive walkthrough

The [Scoring Walkthrough](scoring-walkthrough.qmd) animates this — maximum merge, range masking and the regulatory floors, step by step from input rasters.
:::

## The rule, in one place

The merge rules are **`msens::merge_sql()`**, **`msens::turtle_sql()`** and **`msens::dps_sql()`** — SQL held in the package rather than inside a notebook. The merge notebook executes them and the unit tests assert them against synthetic fixtures, one per taxon category, so the pipeline and its tests cannot drift apart. Changing a rule in a way that breaks a documented case fails the tests.

## Pipeline

```{mermaid}
%%| label: fig-merge-pipeline
%%| fig-cap: "Merging one taxon's source models into a single surface."
%%| fig-width: 6
flowchart TD
  A["Gather this taxon's models<br>across all source datasets"] --> T{"Sea turtle?"}
  T -->|Yes| M["Multiply DPS extinction-risk<br>surface by suitability"]
  T -->|No| P{"NMFS DPS-listed<br>species?"}
  P -->|Yes| Q["Suitability masked to the<br>extinction-risk footprint;<br>per-cell risk kept beside it"]
  P -->|No| B{"Has an expert range<br>anywhere in the world?"}
  Q --> F
  B -->|Yes| C["Range footprint valued at<br>max(governing er_score,<br>AquaMaps at those cells)"]
  B -->|No| D["Raw AquaMaps footprint<br>(no expert range to constrain it)"]
  M --> E["Critical habitat<br>overrides with a max"]
  C --> F["Persist to the<br>merged surface"]
  D --> F
  E --> F
```

### Step 1 — gather the taxon's models

Every model resolving to the same taxon is collected: an AquaMaps suitability surface, any regulatory habitat or range designation (NMFS, FWS), a BirdLife or IUCN range map, and for sea turtles a SWOT+DPS range.

### Step 2 — the expert range constrains AquaMaps

This is the heart of the merge, and it is not a plain maximum.

Where a taxon **has an expert range**, the merged surface is the **range footprint**, valued at the maximum of the taxon's governing `er_score` and the AquaMaps value *at those cells*:

$$
v_{c} = \max\!\big(\text{er\_score},\; v_{am,c}\big) \qquad c \in \text{range}
$$ {#eq-merge-range}

AquaMaps values **beyond** the range are masked away. An environmental-envelope model will happily predict suitable habitat far outside where a species is known to occur; the expert range is what stops that prediction becoming an assertion of presence.

Where a taxon has **no expert range anywhere**, there is nothing to constrain it against, so its raw AquaMaps footprint is kept as-is.

::: {.callout-important}
## "Has a range" is a **global** question, not a local one

Whether a taxon counts as range-constrained is decided on its range **anywhere in the world**, not on whether that range reaches the study area.

That distinction is the entire mechanism for excluding species whose expert range lies wholly outside US waters. Such a species has an expert range, so it takes the range-constrained path; its range ∩ US is empty, so it contributes **no US cells**. Deciding instead on "has a range *in the US*" would push those same species onto the unconstrained AquaMaps path and re-admit roughly **750** of them — river dolphins such as *Sotalia guianensis* among them — on the strength of edge-of-range model artifacts that expert assessment places nowhere near the US.

This was a v6 fix, lost in the v8 rewrite and since restored. It is now a named regression test.
:::

### Step 3 — two surfaces, not one

```{r}
#| label: two-surfaces
#| output: asis

if (mdl_key_style) {
  cat(
"Since v8 the merge emits **two** surfaces from the same rules, because they answer different
questions:

- a **global** surface (expert range ∪ the taxon's whole AquaMaps footprint) — the honest
  whole-range merged model, which is what the species maps draw;
- a **US scoring** surface — the range-constrained, US-clipped result described above, which is
  what the scores are computed from.

Earlier releases produced only the second, so a species map showed the analysis extent rather
than the species. Keeping both means the picture is not silently trimmed to the study area.\n\n",
sep = "")
}
```

### Step 4 — sea turtles multiply instead of taking a maximum

For the six sea turtle species, extinction risk **varies across the range**: NOAA Fisheries lists Distinct Population Segments separately, so the same species can be Endangered in one region and Threatened in another. A single scalar weight cannot express that.

So the turtle merge multiplies the DPS extinction-risk surface by the suitability surface:

$$
v_{c} = \max\!\Big(1,\; \mathrm{round}\Big(\frac{v_{er,c} \times v_{suit,c}}{100}\Big)\Big)
$$ {#eq-merge-turtle}

Critical-habitat designations then override with a maximum, so a designated cell is never scored below its designation.

Because risk is already inside the merged values, turtles receive `er_score = 100` at scoring time as a pass-through multiplier (flagged `is_er_spatial`), which cancels in the division by 100 rather than double-counting.

### Step 5 — NMFS DPS-listed species: the model stays the distribution, risk varies by cell

A second group of species carries extinction risk that varies across the range without being sea turtles: those whose ESA *listed entity* is a Distinct Population Segment, an ESU or a subspecies rather than the whole species — the humpback whale, the Southern Resident killer whale, Cook Inlet beluga, the salmon and steelhead ESUs, both sturgeons, and the others listed under *NMFS Distinct Population Segments* in @sec-data-sources. For them the pipeline builds a per-cell extinction-risk surface (`dps_nmfs`): each listed entity's designated critical habitat at that entity's status, and the species' baseline everywhere else in its IUCN range (the MMPA floor of 20 for a marine mammal; the IUCN category otherwise).

These species do **not** use the turtle multiplication. Their merged value is the **distribution** — the suitability surface (AquaX, or AquaMaps where AquaX did not model the species) masked to the extinction-risk footprint, exactly the range mask every other taxon receives, with a cell that no suitability model covers valued at its risk, as in the plain rule — and the per-cell risk is carried **beside** the value rather than inside it:

$$
v_{c} = \begin{cases} v_{suit,c} & \text{where a suitability model covers } c \\ v_{er,c} & \text{otherwise} \end{cases}
\qquad\qquad er_{c} = v_{er,c}
$$ {#eq-merge-dps}

Scoring then multiplies the two (@sec-scoring), so the score is the same extinction-risk-weighted suitability the turtle rule produces — computed one step later. What changes is what the merged model *shows*: under the turtle rule the humpback, whose risk is the MMPA baseline across 99.8 % of its range, collapsed to a near-flat minimum everywhere outside critical habitat — the map drew the *weight*, not the whale. Now the merged model is the whale's habitat suitability, and its risk is a separate layer (*NMFS DPS extinction risk* in the species app).

Critical-habitat datasets are not an input for these species: `dps_nmfs` already carries every designation at its own entity's status, whereas the generic critical-habitat layers carry the species-level status (Endangered for the humpback) and would override the Threatened Mexico DPS habitat back to Endangered.

::: {.callout-note}
## Three merge strategies

- **Most taxa** — range-constrained maximum, then a **scalar** extinction-risk weight at scoring time.
- **Sea turtles** — an extinction-risk **raster** multiplied into the merged model, so risk varies spatially within the species.
- **NMFS DPS-listed species** — the merged model is the suitability; an extinction-risk **raster** is kept beside it and multiplied in at scoring time.
:::

## Where extinction risk enters

The governing `er_score` (@sec-extinction-risk) is applied **at merge time**, valuing the range cells — the point at which a species' presence in a cell should carry its governing risk rather than whatever code the contributing dataset happened to record. Applying it later, at scoring, would lose that: the merged surface would carry source-specific codes that may be stale or out of jurisdiction. The NMFS DPS-listed species are the deliberate exception: their risk is a per-cell surface built from the listings themselves, so it is carried beside the merged value and applied at scoring, where the per-cell weight replaces the scalar one.

## The statutory floors

The MMPA (20) and MBTA (10) floors are part of the governing `er_score`, not a separate pass over the cells. A marine mammal's range cells therefore carry at least 20 wherever it is present, and a CFR 50 §10.13 migratory bird at least 10 — see @sec-extinction-risk for how each flag is assigned, and why the MBTA one is a species list rather than "all birds".

## Persisted output

```{r}
#| label: merge-outputs
#| output: asis

idf <- doc_manifest()$id_field
cat("The merged result is stored as:\n\n",
"- **`model`** — one row per merged model, keyed by `", idf, "`.\n",
"- **`model_cell`** — one row per model × cell, holding the merged value in `val`.\n\n", sep = "")
if (mdl_key_style) {
  cat(
"`mdl_key` is a **stable** identifier of the form `{dataset}|{species}` — `am|Fis-29291`,
`ms_merge|WORMS:137209`. It replaced an auto-increment id that renumbered on every rebuild,
which broke every saved link to a model.\n\n", sep = "")
}
```

## Validity after merging

A taxon is flagged valid once merging has run, on the criteria in @sec-taxonomy — chiefly that it resolved to a taxon, is marine and not extinct, and has at least one merged cell in the study area. Program-Area overlap is **not** among them from v7 onward; it is a spatial question answered later.

```{r}
#| label: merge-counts
#| output: asis

cat("In ", doc_ver(), " this yields **", doc_stat("valid_species"),
    " valid species** across the study area", sep = "")
if (doc_can("programareas")) cat(", of which **", doc_stat("species_program_areas"),
    "** contribute to scores within the ", doc_stat("n_program_areas"),
    " BOEM Program Areas of the current program cycle", sep = "")
cat(".\n\n")
```

## Guarding the rules

Every merge category has a synthetic fixture in the `msens` test suite asserting its exact expected output: range-only, both-masked, range-outside-the-US excluded, AquaMaps-only single-model, AquaMaps-only multi-model (which must **not** deduplicate), and the multiplicative turtle case.

This is deliberate. Whole-pipeline checks — the version-equivalence gate on Program-Area scores, and the content-hash checkpoints — catch *aggregate* drift, but they hide rule-level breakage: the ~750 wrongly-included species above moved aggregate scores almost not at all while being scientifically wrong. A category with no fixture is a category that can break silently.
