# Releases {#sec-releases}

```{r}
#| include: false
library(dplyr)
rel <- doc_releases()
this <- doc_version_row()
```

This book is published **once per release**. The page you are reading documents **`r doc_ver()`**, and every figure in it comes from that release's own published tables — so a v3 page reports v3's numbers, not the newest ones. Use the version switcher in the sidebar to move between builds.

This page is the exception: it reads *all* releases, because the comparison is exactly what a single-version page cannot show.

```{r}
#| label: this-release
#| output: asis

if (nrow(this)) {
  st <- c(released = "the promoted release", prerelease = "a **pre-release**",
          retired = "**retired** — superseded, but still published and still readable")
  cat("::: {.callout-note}\n## ", doc_ver(), " — ", this$title, "\n\nReleased ",
      as.character(this$released), ". This is ", st[[this$status]], ".\n",
      if (identical(this$status, "prerelease"))
        "\nA pre-release is reachable only by naming it explicitly; it is never what a reader gets by default.\n" else "",
      ":::\n", sep = "")
}
```

## Every release at a glance

```{r}
#| label: tbl-releases
#| tbl-cap: "Published releases. Species counts are read from each release's own `taxon` table and are **not** comparable across releases, because the definition of a valid species changed at v3 and again at v7 (see below)."

knitr::kable(
  data.frame(
    Version    = ifelse(rel$ver == doc_ver(), paste0("**", rel$ver, "**"), rel$ver),
    Released   = rel$released,
    Status     = rel$status,
    Title      = rel$title,
    Datasets   = n_fmt(rel$n_datasets),
    `Valid species` = n_fmt(rel$valid_species),
    `In Program Areas` = ifelse(is.na(rel$species_pra), "—", n_fmt(rel$species_pra)),
    Grid       = paste0("`", rel$grid, "`"),
    check.names = FALSE, row.names = NULL),
  align = c("l", "l", "l", "l", "r", "r", "r", "l"))
```

::: {.callout-warning}
## Why the species count is not a trend line

The count moves for three different reasons, only one of which is "more data":

- **v3** tightened the taxonomy — resolving every source name to a canonical identity, dropping non-marine and extinct taxa — which cut the count roughly in half.
- **v6** removed species whose expert range lies entirely outside US waters, whose apparent US presence was an artifact of environmental-envelope prediction.
- **v7** stopped requiring Program-Area overlap for validity. Nothing was added; the definition changed, and the count went up accordingly.

Read the column as "what this release considered valid", never as a measurement of biodiversity that changed between releases.
:::

```{r}
#| label: tbl-releases-tech
#| tbl-cap: "What each release published, read from its manifest."

knitr::kable(
  data.frame(
    Version       = rel$ver,
    `Model id`    = paste0("`", rel$id_field, "`"),
    Metrics       = n_fmt(rel$n_metrics),
    `Spatial units` = rel$units,
    check.names = FALSE, row.names = NULL),
  align = c("l", "l", "r", "l"))
```

## History

### v8 — Marine Atlas

A ground-up rebuild as a global **Marine Atlas**.

- **Whole-range global ingest.** Every species is modeled across its *entire* global range on a shared global 0.05° ocean grid, with no land mask and no Program-Area gate, so a distribution map shows the species rather than the analysis extent. "How marine" becomes a *derived* measurement rather than a filter applied up front.
- **Seabird de-duplication.** A BirdLife ↔ WoRMS crosswalk merges each bird's BirdLife, USFWS and AquaMaps inputs into a *single* taxon; v7 split some seabirds across two (Marbled Murrelet, Brown Pelican among them), double-counting them. Correcting it lowers seabird-heavy Program Areas in Alaska and the Pacific by several points, with overall Program-Area correlation to v7 of 0.94.
- **Taxonomy-driven categories.** Categories come from WoRMS class and phylum rather than keyword matching, which removed the catch-all `other` bucket and surfaced `primary_producer`. Non-turtle reptiles and amphibians are excluded from scoring.
- **Original *and* interpolated distributions.** Each source model is published in two representations — the native-resolution original and the 0.05° interpolated surface used in scoring — catalogued in STAC and toggleable in the species app.
- **Partitioned Parquet on S3.** Releases are published as partitioned, content-addressed Parquet with COGs, PMTiles and a STAC catalog, replacing per-dataset archives and a monolithic database file.
- **A stable model identifier.** `mdl_key` (`am|Fis-29291`, `ms_merge|WORMS:137209`) replaces the auto-increment `mdl_seq`, which renumbered on every rebuild and so broke every saved link to a model.
- **A different grid.** v8 moves from a regional grid on 0–360 longitudes to a global one on −180 to 180, so **`cell_id` denotes a different place** than it did in v1–v7.

### v7 — validity decoupled from Program Areas

Earlier releases silently required a species to overlap a current BOEM Program Area to count as valid, which hid biodiversity outside the program cycle's footprint. Validity now means a taxonomically valid species with a modeled distribution anywhere in the study area, restoring the set from 9,424 to 16,153 — of which 9,230 fall within the 20 Program Areas. Program-Area membership became a spatial query rather than a species-level filter. Scores were recomputed over the full set with negligible effect: 18 of 20 Program Areas unchanged, and only Gulf of America GAB (−0.04) and GAA (−0.00) shifted at all.

### v6 — expert ranges outside the US EEZ

Excluded 371 species (343 fish, 20 invertebrates, 6 corals, 2 mammals) whose IUCN expert ranges fall entirely outside the US EEZ; their AquaMaps predictions in US waters were edge-of-range artifacts with near-zero suitability. Also ingested three missing marine-mammal IUCN ranges (*Phoca vitulina*, *Trichechus manatus*, *Pusa hispida*) from a more complete download. Effect on Program-Area scores was negligible.

This exclusion was later lost in the v8 rewrite and restored; it now has a named regression test, because ~750 species were affected while aggregate scores barely moved — precisely the kind of error a whole-pipeline check cannot see (@sec-model-merging).

### v5 — MMPA floor for FWS-managed marine mammals

Fixed the MMPA spatial floor for marine mammals managed by FWS rather than NMFS — walrus, polar bear, sea otter, manatee, dugong — which had been missing the protection flag because it was derived from the NMFS Species Directory alone. Later releases assign the flag by taxonomy instead, so no marine mammal can be missed (@sec-extinction-risk).

### v4b — the multiplicative turtle merge

Introduced the second modeling approach, for sea turtles only: instead of taking a maximum and applying a scalar extinction-risk weight at scoring time, the SWOT+DPS extinction-risk **raster** is multiplied into the suitability surface during merging, so risk varies spatially within a species — matching the fact that NOAA Fisheries lists Distinct Population Segments separately. Critical habitat overrides with a maximum. Affected taxa are flagged `is_er_spatial` and pass `er_score = 100` through at scoring so the risk is not counted twice.

This was the point at which the MST had **two** merge strategies rather than one, and it went undocumented for two releases (see @sec-model-merging).

### v4 — SWOT sea turtle ranges

Replaced the overly extensive IUCN sea turtle range maps with the more realistic [State of the World's Sea Turtles](http://seamap.env.duke.edu/swot) global distributions, and applied NOAA Distinct Population Segment polygons carrying ESA listing status to distinguish Endangered from Threatened populations. The result is a more restrictive distribution dominated by ESA risk values.

### v3 — merged models and extinction risk

Program Areas for the National OCS Plan, and ecoregional rescaling. Source datasets merged from continuous suitability models through to binary range maps, with masking where a global IUCN range exists. Introduced species sensitivity weighting from national ESA or global IUCN extinction risk, together with MMPA and MBTA protection. Sea turtles were split out of `reptile` into their own category.

### v2 — Program Areas

Transition to the 2026 Program Areas. Individual model layers, a model-merging pipeline, and cloud-optimized GeoTIFF storage.

### v1 — Planning Areas

Planning Areas across the US EEZ, AquaMaps-based species distributions, flower plots for multi-taxa visualization, and species tables with conservation status. See @sec-release-v1 for the full v1 methodology as delivered in the 2025 final report.

## Reading an older release

Every release remains published and readable:

- **Documentation** — this book, at `/{version}/`; the sidebar switcher moves between builds and marks any version whose docs are not built.
- **Applications** — [`/scores/?ver=v3`](https://app.marinesensitivity.org/scores/?ver=v3) and [`/species/?ver=v3`](https://app.marinesensitivity.org/species/?ver=v3), substituting any published version. The per-version application forks are retired; their old URLs 301 to the equivalent `?ver=`, so links printed in earlier reports still resolve to the release they always showed.
- **Data** — `marine-atlas/{version}/tables/*.parquet` on S3, readable directly by DuckDB (including in a browser) without downloading a database.

```{r}
#| label: promotion
#| output: asis

lat <- tryCatch(msens::atlas_latest(), error = function(e) NA_character_)
if (!is.na(lat)) cat(
  "**Promotion is deliberate and separate from publishing.** Writing data does not make a release ",
  "current: the pointer that applications follow by default is only rewritten under an explicit ",
  "flag, and may only name a released — never a pre-release — version. It currently names **",
  lat, "**.\n\n", sep = "")
```
