# Releases {#sec-releases}

```{r}
#| include: false
library(dplyr)
rel <- doc_releases()
this <- doc_version_row()
```

This book is published **once per release**. The page you are reading documents **`r doc_ver()`**, and every figure in it comes from that release's own published tables — so a v3 page reports v3's numbers, not the newest ones. Use the version switcher in the sidebar to move between builds.

This page is the exception: it reads *all* releases, because the comparison is exactly what a single-version page cannot show.

```{r}
#| label: this-release
#| output: asis

if (nrow(this)) {
  st <- c(released = "the promoted release", prerelease = "a **pre-release**",
          retired = "**retired** — superseded, but still published and still readable")
  cat("::: {.callout-note}\n## ", doc_ver(), " — ", this$title, "\n\nReleased ",
      as.character(this$released), ". This is ", st[[this$status]], ".\n",
      if (identical(this$status, "prerelease"))
        "\nA pre-release is reachable only by naming it explicitly; it is never what a reader gets by default.\n" else "",
      ":::\n", sep = "")
}
```

## Every release at a glance

```{r}
#| label: tbl-releases
#| tbl-cap: "Published releases. Species counts are read from each release's own `taxon` table and are **not** comparable across releases, because the definition of a valid species changed at v3 and again at v7 (see below)."

knitr::kable(
  data.frame(
    Version    = ifelse(rel$ver == doc_ver(), paste0("**", rel$ver, "**"), rel$ver),
    Released   = rel$released,
    Status     = rel$status,
    Title      = rel$title,
    Datasets   = n_fmt(rel$n_datasets),
    `Valid species` = n_fmt(rel$valid_species),
    `In Program Areas` = ifelse(is.na(rel$species_pra), "—", n_fmt(rel$species_pra)),
    Grid       = paste0("`", rel$grid, "`"),
    check.names = FALSE, row.names = NULL),
  align = c("l", "l", "l", "l", "r", "r", "r", "l"))
```

::: {.callout-warning}
## Why the species count is not a trend line

The count moves for three different reasons, only one of which is "more data":

- **v3** tightened the taxonomy — resolving every source name to a canonical identity, dropping non-marine and extinct taxa — which cut the count roughly in half.
- **v6** removed species whose expert range lies entirely outside US waters, whose apparent US presence was an artifact of environmental-envelope prediction.
- **v7** stopped requiring Program-Area overlap for validity. Nothing was added; the definition changed, and the count went up accordingly.

Read the column as "what this release considered valid", never as a measurement of biodiversity that changed between releases.
:::

```{r}
#| label: tbl-releases-tech
#| tbl-cap: "What each release published, read from its manifest."

knitr::kable(
  data.frame(
    Version       = rel$ver,
    `Model id`    = paste0("`", rel$id_field, "`"),
    Metrics       = n_fmt(rel$n_metrics),
    `Spatial units` = rel$units,
    check.names = FALSE, row.names = NULL),
  align = c("l", "l", "r", "l"))
```

## History

Each release is summarized under the same four headings, so two versions can be read line for line. The same summary heads that version's own landing page.

- **Datasets** — inputs added or changed, and the role each plays (contributing a value, masking a range, or carrying extinction risk spatially).
- **Methods** — the scientific rules: how models merge, how extinction risk is scored, which statutory floors apply.
- **Scope** — what counts and where: the definition of a valid species, the spatial units, the study area, which taxa are eligible.
- **Technology** — formats, identifiers, serving and applications.

::: {.callout-warning}
## Scope changes move the count without any data changing

The distinction between **Methods** and **Scope** is the one to watch. v3 tightened the taxonomy, v6 excluded ranges outside the EEZ, and v7 decoupled validity from Program Areas — each moved the headline species count substantially, and none of them added or removed a single input. Reading those jumps as growth or loss is the single easiest mistake to make with this table.
:::

```{r}
#| label: history
#| output: asis

# Rendered from the SAME curated bullets + computed deltas the landing pages use,
# so the two cannot disagree about what a release changed.
for (v in rel$ver) doc_changes_callout(v)
```

## Reading an older release

Every release remains published and readable:

- **Documentation** — this book, at `/{version}/`; the sidebar switcher moves between builds and marks any version whose docs are not built.
- **Applications** — [`/scores/?ver=v3`](https://app.marinesensitivity.org/scores/?ver=v3) and [`/species/?ver=v3`](https://app.marinesensitivity.org/species/?ver=v3), substituting any published version. The per-version application forks are retired; their old URLs 301 to the equivalent `?ver=`, so links printed in earlier reports still resolve to the release they always showed.
- **Data** — `marine-atlas/{version}/tables/*.parquet` on S3, readable directly by DuckDB (including in a browser) without downloading a database.

```{r}
#| label: promotion
#| output: asis

lat <- tryCatch(msens::atlas_latest(), error = function(e) NA_character_)
if (!is.na(lat)) cat(
  "**Promotion is deliberate and separate from publishing.** Writing data does not make a release ",
  "current: the pointer that applications follow by default is only rewritten under an explicit ",
  "flag, and may only name a released — never a pre-release — version. It currently names **",
  lat, "**.\n\n", sep = "")
```
