# Taxonomy {#sec-taxonomy}

```{r}
#| include: false
library(dplyr)
cats <- doc_sp_cats()
```

Integrating distribution data from several providers requires resolving taxonomic identity across naming systems, so that models of the same species merge rather than being counted as different species. The MST uses a multi-authority matching pipeline in which every source name resolves to one canonical identifier.

## Two taxonomic namespaces

Taxa resolve into **one of two** authorities, not one:

- **Birds → BirdLife (`BOTW:{sisid}`)**. WoRMS coverage of birds is poor and its `isMarine` flag is unreliable for seabirds, so birds are *not* resolved through WoRMS.
- **Everything else → WoRMS (`WORMS:{aphia_id}`)**, resolved from the scientific name via the ITIS↔WoRMS crosswalk, then exact WoRMS matching, then the WoRMS REST API for the remainder.

A merged model is keyed by that identity — `ms_merge|WORMS:137209`, `ms_merge|BOTW:22694927` — which is also why species must be counted on the **(authority, id) pair**: the two namespaces number independently, so counting bare ids collides birds with invertebrates.

## Taxonomic authorities

| Authority | Role | Key |
|:----------|:-----|:----|
| [WoRMS](https://www.marinespecies.org) | World Register of Marine Species — canonical identity for all non-bird marine taxa, and the source of the class/phylum used to assign a species category | `worms_id` (AphiaID) |
| [BirdLife BOTW](https://www.birdlife.org) | Birds of the World — canonical identity and ranges for birds | `botw_id` (SIS ID) |
| [GBIF](https://www.gbif.org) | Backbone taxonomy, used as a crosswalk into WoRMS | `gbif_id` |
| [ITIS](https://www.itis.gov) | US federal standard; the first match attempted where a source supplies a TSN | `itis_tsn` |
| [IUCN Red List](https://www.iucnredlist.org) | Conservation status, and expert range maps | `iucn_id` |

: Taxonomic authorities used for name resolution. {#tbl-tax-authorities}

At each stage deprecated names resolve to their accepted synonyms, with the original source name kept alongside the accepted one.

## Species categories

```{r}
#| label: tbl-species-categories
#| tbl-cap: !expr paste0("Species categories in ", doc_ver(), ". *Scored* is read from the metric registry — a category is scored only where the release published an `extrisk_{cat}` metric for it. *Valid* counts the taxa eligible for scoring.")

knitr::kable(
  data.frame(
    Category = paste0("`", cats$sp_cat, "`"),
    Taxa     = formatC(cats$n_taxa,  format = "d", big.mark = ","),
    Valid    = formatC(cats$n_valid, format = "d", big.mark = ","),
    Scored   = ifelse(cats$scored, "Yes", "**No**"),
    check.names = FALSE, row.names = NULL),
  align = c("l", "r", "r", "c"))
```

```{r}
#| label: cats-notes
#| output: asis

unscored <- cats$sp_cat[!cats$scored]
if (length(unscored)) cat(
  "Categories present in the taxonomy but **not scored** in ", doc_ver(), ": ",
  paste0("`", unscored, "`", collapse = ", "),
  ". They are carried in the taxon table — the identity is real — but no metric ",
  "aggregates them, so they contribute nothing to any score.\n\n", sep = "")

if (!"other" %in% cats$sp_cat) cat(
  "There is **no `other` bucket** in this release: every taxon lands in a named ",
  "category assigned from its WoRMS class/phylum, or from BirdLife for birds.\n\n", sep = "")
```

```{r}
#| label: cats-callout
#| output: asis

# What is true about categories differs BY RELEASE, so this is assembled from the
# release's own category set rather than written once and printed everywhere.
# Static v8 prose here put `primary_producer` on a v3 page, where the category
# does not exist and `other` still does.
have <- cats$sp_cat
cat("::: {.callout-note}
## How categories are assigned

")
if ("primary_producer" %in% have) cat(
  "The category comes from **taxonomy**, not from the dataset a model arrived in: WoRMS class ",
  "and phylum for non-birds, BirdLife membership for birds.

", sep = "")
if ("turtle" %in% have) cat(
  "- **Sea turtles are their own category (`turtle`)**, separated from `reptile` so that the ",
  "multiplicative turtle merge and the turtle metric have somewhere to live ",
  "(see @sec-model-merging). Non-turtle reptiles — sea snakes, marine iguanas — remain in the ",
  "taxonomy but are not scored.
", sep = "")
if ("primary_producer" %in% have) cat(
  "- **`primary_producer`** (macroalgae, seagrasses and other photosynthesisers) is a category in ",
  "its own right. Taxa missing from the bulk WoRMS download — chiefly AlgaeBase names — are ",
  "fetched individually from the WoRMS API and cached, rather than being swept into a bucket.
",
  sep = "")
if ("other" %in% have) cat(
  "- **`other`** collects taxa this release's assignment could not place. It is a symptom rather ",
  "than a category, and later releases removed the need for it by assigning from the taxonomic ",
  "hierarchy instead — see [Releases](releases.qmd).
", sep = "")
cat(":::

")
```

## Marine relevance {#sec-is-marine}

Not every taxon in the source data belongs in a *marine* sensitivity analysis, and the obvious test does not work.

**WoRMS `isMarine` fails for birds.** Birds are resolved through BirdLife, so most have no WoRMS record at all, and those that do are inconsistently flagged; relying on it would drop nearly every seabird. Earlier releases sidestepped this with a purely **spatial** test — a taxon counted as marine if its distribution overlapped the study area — which admits any coastal or inland bird whose range happens to reach the coast.

```{r}
#| label: marine-cull
#| output: asis

if (!is.na(.doc_keys()$marine)) {
  cat(
"This release applies an explicit **marine-relevance cull** to birds, combining two tests
with a curation list:

1. the species belongs to a **marine or coastal family** (an explicit allow-list of families), **and**
2. its **whole global range** is at least 5% marine, measured as the share of its range cells
   that are ocean.

Curated CSVs add and remove individual species where the rule gets it wrong. The measurement is
taken over the **entire** range rather than the US portion, so a bird is judged on what it is,
not on where the study area happens to cut it. Sub-pixel islands are a known confounder — at
0.05° a small island can make a strictly terrestrial species look partly marine — which is what
the family allow-list guards against.\n\n", sep = "")
} else {
  cat(
"This release predates the explicit marine-relevance cull. Marine status comes from WoRMS
`isMarine`, with birds admitted through BirdLife instead, so the bird counts on this page are
broader than in later releases.\n\n", sep = "")
}
```

## From source taxa to valid species

```{r}
#| label: tbl-filter-funnel
#| tbl-cap: !expr paste0("Validity gates applied cumulatively to ", doc_ver(), "'s published `taxon` table. The final row applies the release's own validity flag to the whole table — it is the authoritative count, and not simply the conjunction of the diagnostic gates above it.")

f <- doc_funnel()
knitr::kable(
  data.frame(
    `Filter step`    = f$step,
    Removed          = ifelse(f$removed <= 0, "—", paste0("−", n_fmt(f$removed))),
    `Taxa remaining` = n_fmt(f$remaining),
    check.names = FALSE, row.names = NULL),
  align = c("l", "r", "r"))
```

This yields **`r doc_stat("valid_species")` valid species** across the study area.

```{r}
#| label: pra-subset
#| output: asis

if (doc_can("programareas")) cat(
  "Of those, **", doc_stat("species_program_areas"), "** have modeled distributions inside the **",
  doc_stat("n_program_areas"), " BOEM Program Areas** of the current program cycle. That is a ",
  "**spatial** result, obtained by asking the zone tables which species occur there — not a ",
  "property of the validity flag.\n\n", sep = "")
```

::: {.callout-important}
## Validity is not the same question as "occurs in a Program Area"

Through v6 the validity flag silently *required* overlap with a BOEM Program Area, conflating "this is a real, mapped marine species" with "this species happens to fall inside the areas of the current program cycle" — and so hiding biodiversity outside that cycle. From **v7** the two are separate: validity means the taxon has a modeled distribution in the study area, and Program-Area membership is a zone query answered afterwards.

This is why the headline count is **not comparable across releases**: it rose sharply at v7 without a single species being added, because the definition changed. Per-release counts are on the [Releases](releases.qmd) page.
:::

## Data quality

### Duplicate resolution

Several datasets may supply a model for the same species. Resolving every source name to one canonical identifier is what makes those models merge — taking the maximum per cell — instead of appearing as separate species and counting the taxon more than once in every score.

### Synonym handling

Names change as taxa are reclassified. The pipeline resolves through `acceptedNameUsageID` in each authority, keeps the original source name alongside the accepted one, and updates deprecated identifiers to their current accepted form.

## Implementation

Name resolution is `msens::match_taxa()`, which runs the identifier cascade and returns the unified taxon table. Category assignment, the marine cull and the validity flags are applied in the `merge_taxon` workflow (see @sec-workflows).
